---
title: "Intro to MLM: Practical 3"
author: Patricio Troncoso and Ana Morales-Gómez
date: "June 2022"
output: 
  html_document:
    code_download: yes
    highlighter: null
    theme: cosmo
    toc: yes
    toc_depth: 4
    toc_float: yes
    fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


For this practical, we will use data from a sample of the Teaching dataset of the Health Survey for England (2003-2005), which is
available [**here**](https://beta.ukdataservice.ac.uk/datacatalogue/studies/study?id=6765).

We will be fitting a series of binary logistic multilevel models for the probability of being overweight or obese.

The traditional measure used to determine whether an adult is overweight is the Body-Mass Index (BMI).

For an adult to be considered overweight, their BMI must be between 25 and 30. Meanwhile values over 30 are considered as "obese".

<br>

***

# Typical workflow setup and data preparation

## 1.1. Define a working directory

You can use any directory in your computer. As in the example below:

```{r, eval=F}
setwd("C:/myfolder")
```

Remember to download the data to the folder you will define as working directory, as this makes matters easier.

## 1.2. Load packages

You can always load packages later on, but it is a good practice to load packages at the beginning of the session on the top of your script or R markdown file.

In this practical, we will use the packages `haven`, `lme4` and `ggplot2`. Remember that if you haven't installed them before, you need to do so before you call the `library` function:

```{}
install.packages("haven")
install.packages("lme4")
install.packages("ggplot2")
install.packages("dplyr")
```

Then you load them as such:

```{r, warning=F, message=F}
library(haven)
library(lme4)
library(ggplot2)
library(dplyr)
```

## 1.3. Read in data 

You can download the data from  the [**UKDS website**](https://beta.ukdataservice.ac.uk/datacatalogue/studies/study?id=6765). There are four files in different formats. We will be using the Stata dataset named: "hse_data_for_workshop.dta".

To read this dataset into R, we need to use the package `haven`:


```{r, warning=F, message=F}
hse <- read_dta("https://github.com/A-mora/MLM_summer-school/raw/main/data/hse_data_for_workshop.dta")
```

<br>

***

## Select variables to use

The dataset has many different variables, but we're not going to be using all of them, so for easier work, we will select the following: `year`, `pserial`, `area`, `bmival`, `sex`, `age` and `ethnic`.

`year` is the year where the measures were taken

`pserial` is the anonymised person identifier

`area` is the anonymised postcode of the person

`bmival` is the valid measure of BMI

`sex` coded 0 for Men and 1 for Women

`age` from 0 to 99

`ethnic` coded 1 for White; 
               2 for Mixed; 
               3 for Asian or Asian British; 
               4 for Black or Black British and 
               5 for Chinese or any other group

For a detailed description of these variables, have a look in the documentation downloaded alongside with the data. 
**Spoiler alert:** You will also find a complete set of materials for a workshop on MLM.

To select variables, we (unsurprisingly) need to use the function `select` of the `dplyr` package:

```{r, warning=F, message=F}

hse2 <- select(hse, year, pserial, area, bmival, sex, age, ethnic, bmival) 

```

<br>

***
# Preliminary task

We need to dichotomise our dependent variable `bmival`. As we mentioned before, the categories "overweight" and "obese" are for those with a BMI over 25 and 30, respectively, so we need to do the following to our continuous BMI measure:

```{r, warning=F, message=F}
hse2$bmi_bin <- ifelse(hse2$bmival>=25, 1, 0)
```

To check this is done correctly, you can 


# Task 1: Fit an empty model

We will start by fitting an empty multilevel model for the propensity of being overweight in people nested in areas.

```{r, warning=F, message=F}
empty <- glmer(bmi_bin ~ 1 + (1|area), data = hse2, family = binomial("logit"))

summary(empty)
```

### Questions:

1.1 What is the overall probability of being overweight?

Remember this formula:

$$logit(p_i)=\log\left(\frac{p_{i}}{1-p_{i}}\right)$$
To get the odds from the estimated coefficients, you use the exponent function:

$$exp(\beta)=odds$$

Then, to convert to probability, you can do this:

$$p = \frac{odds}{1 + odds}$$
```{r, warning=F, message=F}

odds <- exp(empty@beta) # this is to retrieve the intercept from the model

prob <- odds/(1 + odds) # to estimate the overall probability

```


1.2 What is the VPC for this model?

$$VPC = \frac{\sigma_u^2}{\sigma_u^2 + \sigma_e^2*}$$
where $\sigma_e^2*= \frac{\pi^2}{3} \approx 3.29$

```{r, warning=F, message=F}
lev2var <- as.numeric(VarCorr(empty)) # this is to retrieve the level 2 variance from the model

(lev2var)/(lev2var+3.29) # VPC
```

<br>

***


# Task 2: Is the MLM better than a single-level model?

To assess the statistical significance of the MLM, we need to compare it to a single-level model. To fit a single-level model, we type:

```{r, warning=F, message=F}

single <- glm(bmi_bin ~ 1 , data = hse2, family = binomial("logit"))

summary(single)

```

These results are not very interesting in themselves, so we move on extract the loglikelihood and compare to the MLM.

```{r, warning=F, message=F}

L1 <- as.numeric(logLik(single)) # store this as numerical to re-use
L2 <- as.numeric(logLik(empty))

(D <- 2*(L2-L1)) # this is the deviance, we put it within brackets to print it immediately

```

### Question

3.1. Is the addition of the area level statistically significant?

3.2. What does this mean in practice?

***

# Task 3: Random intercepts model

```{r, warning=F, message=F}
rand_int <- glmer(bmi_bin ~ 1 + (1|area), data = hse2, family = binomial("logit"))

summary(empty)
```
